---
title: "Title"
author: "Jefferson Pastuña"
date: "2024-04-26"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Introduction

Introduction...

## Before to start

Before to start...

## Notame workflow

As a first step...

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame", ref = "v0.3.1")

# Notame library call
library(notame)

# Dependency packages installation
install_dependencies

```

Then, a main path and a log system was added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "../Ilex_guayusa_methylxanthine/"
# Log system
init_log(log_file = paste0(ppath, "Result/Data_log.txt"))

```

Next...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data <- read_from_excel(file = "Data/Target_Mateo_to_R.xlsx", sheet = 2, 
                        corner_row = 9, corner_column = "F", 
                        split_by = c("Column", "Ion Mode"))

```

Once the data is read, the next step was...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")


```

## Preprocessing

The first step is...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode <- modes$RP_POS
# Change 0 value to NA
mode <- mark_nas(mode, value = 0)

```

Then...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
#mode <- flag_detection(mode, qc_limit = 10/14, group_limit = 2/3)

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after low detection algorithm
#visualizations(mode,
#               prefix = paste0(ppath,
#                               "Result/notame_Result/HS_GCMS/Figure/",
#                               "Low_Detection")
#               )

```

The next step...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected <- correct_drift(mode)
corrected <- correct_drift(corrected)
corrected <- correct_drift(corrected)
# Flag low quality features
#corrected <- flag_quality(corrected, condition = "RSD_r < 0.8 & D_ratio_r < 1.6")

```

Then we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp <- plot_sample_boxplots(corrected,
                                order_by = "NewGroup",
                                fill_by = "NewGroup")
# PCA
corr_pca <- plot_pca(corrected,
#                     label = "Sample_ID",
                     center = TRUE,
                     shape = "NewGroup",
                     color = "NewGroup"                     )
# Package to plots visualization in a same windows
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("thomasp85/patchwork")
library(patchwork)
# Plot
corr_pca + corr_bp

```

Batch correction...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Batch correction
corrected_batch <- normalize_batches(corrected, batch = "Batch", group = "QC", ref_label = "QC")

```

Then we can visualize the data after batch correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
batch_bp <- plot_sample_boxplots(corrected_batch,
                                order_by = "NewGroup",
                                fill_by = "NewGroup")
# PCA
batch_pca <- plot_pca(corrected_batch,
#                      label = "Sample_ID",
                      center = TRUE,
                      shape = "NewGroup",
                      color = "NewGroup")
# Plot
batch_pca + batch_bp

```

Contaminant...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Removal of contaminants
#corrected_no_blank <- flag_contaminants(corrected_batch,
#                                        blank_col = "Group",
#                                        blank_label = "Blank",
#                                        flag_thresh = 1.9,
#                                        flag_label = "Contaminant")
# Removal blank group from dataset
#corrected_no_blank <- corrected_no_blank[, corrected_no_blank$Group != "Blank"]

```

Data visualization...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Boxplot
#no_blank_bp <- plot_sample_boxplots(corrected_no_blank,
#                                 order_by = "Factor",
#                                 fill_by = "Factor")
# PCA
#no_blank_pca <- plot_pca(corrected_no_blank,
#                      center = TRUE,
#                      shape = "Factor",
#                      color = "Factor")
# Plot
#no_blank_pca + no_blank_bp

```

The next step...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#clustered <- cluster_features(corrected_no_blank,
#                              rt_window = 2/60,
#                              corr_thresh = 0.95,
#                              d_thresh = 0.8,
#                              #plotting = TRUE,
#                              #prefix = paste0(ppath, "Result/notame_Result/HS_GCMS/Cluster")
#                              )
#compressed <- compress_clusters(clustered)

```

We can visualize data...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
#clust_bp <- plot_sample_boxplots(compressed,
#                                 order_by = "Group",
#                                 fill_by = "Species")
# PCA
#clust_pca <- plot_pca(compressed,
#                      center = TRUE,
#                      shape = "Species",
#                      color = "Species")
# Plot
#clust_pca + clust_bp

```

Finally the data is...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#save(imputed, file = paste0(ppath, "Result/Notame_Result.RData"))

```

# PCA plots

Here read this paper: https://doi.org/10.1007%2Fs11306-016-1030-9

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Impute missing values using random forest
# To clean data
set.seed(615)
imputed <- impute_rf(corrected_batch)
# To all data
imputed <- impute_rf(corrected_batch, all_features = TRUE)
# Exporting to MetaboAnalyst
#write_to_excel(imputed, "Result/Mateo_Data_imputed.xlsx")

```

We can inspect PCA plot after...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp <- plot_sample_boxplots(imputed,
                               order_by = "NewGroup",
                               fill_by = "NewGroup")
# PCA
imp_pca <- plot_pca(imputed,
                    center = TRUE,
                    shape = "NewGroup",
                    color = "NewGroup")
# Plot
imp_pca + imp_bp

```

The next step...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Probabilistic quotient normalization
#pqn_set <- pqn_normalization(imputed,
#                             ref = c("qc", "all"),
#                             method = c("median", "mean"),
#                             all_features = FALSE)

```

We can inspect PCA plot after...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
#pqn_bp <- plot_sample_boxplots(pqn_set,
#                               order_by = "Group",
#                               fill_by = "Species")
# PCA
#pqn_pca <- plot_pca(pqn_set,
#                    center = TRUE,
#                    shape = "Species",
#                    color = "Species")
# Plot
#pqn_pca + pqn_bp

```

Droping flagged features...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Drop QC
no_qc <- drop_qcs(imputed)
# Sub QC data extraction
data_1hm <- no_qc[, no_qc$NewGroup != "Sample"]
# Extract clean data
no_flag <- drop_flagged(data_1hm)
# Extracting feature height table
peak_height <- exprs(no_flag)
# Extracting Phenotipic data
pheno_data <- no_flag@phenoData@data

```

The next step is...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Package to generalised logarithmic transform
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("pmp")
#library(pmp)
# Generalised logarithmic transform
#glog_set <- glog_transformation (df = peak_height,
#                                 classes = pheno_data$Group,
#                                 qc_label = "QC")

```

Preparing data and transposing feature table.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table  <- t(peak_height)
# Centering and Scaling features
ei_pca <- prcomp(transp_table, center = TRUE, scale. = FALSE)

```

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Library to left_join use
library(dplyr)
# PCA scores
scores <- ei_pca$x %>%                   # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(pheno_data )                 # Adding metadata
# PCA plot
ggplot(scores,
       aes(PC1, PC2, shape = Factor, color = Factor)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (23.65 %)"),
         y=guide_axis(title = "PC2 (20.46 %)")) +
  theme_classic()# +
#  theme(legend.text = element_text(face="italic")) +
#  theme(legend.position = c(0.15, 0.75),
#        legend.background = element_rect(fill = "white", color = "black")) +
#  theme(panel.grid = element_blank(), 
#        panel.border = element_rect(fill= "transparent")) +
#  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
#  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
# Save plot
#ggsave('Result/GCMS/GCMS_PCA_To_DQ.pdf', width = 5, height = 4, device='pdf', dpi="print")

```

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings <- ei_pca$rotation %>%           # Extract loadings
  data.frame(Feature_ID = rownames(.))    # New column with feat name

```

Creating an artificial table with Feature name and Compound column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Load a metabolite name table
#metab_name <- readxl::read_excel("Data/GCMS_Metabolites.xlsx", 6)
# Creating a new small table of the annotated compounds
#ei_compouds <- left_join(metab_name, loadings)
# Plotting results
ggplot(loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.2) +
  theme_classic() + 
  geom_point(data = loadings, size = 1) +
  ggrepel::geom_label_repel(data = loadings,
                            aes(label = imputed@featureData@data$Metabolite),
                            box.padding = 0.8,
                            label.padding = 0.3,
                            label.r = 0.3,
                            cex = 3.7) +
  guides(x=guide_axis(title = "PC1 (23.65 %)"),
         y=guide_axis(title = "PC2 (20.46 %)")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray") +
  ggsci::scale_color_aaas()
# Save plot
#ggsave('Result/GCMS/GCMS_Loadings_to_DQ.pdf', width = 5, height = 4, device='pdf', dpi="print")

```

# Heat map plot

ComplexHeatmap package and dependency installation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# ComplexHeatmap package installation
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)

# ColorRamp2 package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("jokergoo/colorRamp2")
library(colorRamp2)

# Cowplot package installation
#install.packages("cowplot")
library(cowplot)

```

Extracting and loaded of identified metabolites abundance.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Metabolite name table without internal standard
#metab_name_hm <- readxl::read_excel("Data/GCMS_Metabolites.xlsx", 6)
# Add identified metabolite to "notame" output
#add_met <- join_fData(no_flag, metab_name_hm)
# Extracting identified metabolite data
#raw_hm <- add_met[!is.na(add_met@featureData@data$Metabolite_name),]
# Extracting feature height table
#hm_height <- exprs(raw_hm)
# Extracting sample information
#hm_pdata <- raw_hm@phenoData@data
# Extracting feature information
#hm_fdata <- raw_hm@featureData@data

```

## First Heatmap plot

Here we extract and plotting heat map of sub QC.

Scaling, row and top heatmap anotation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2500)
# Logarithmic scale
scl_1hm <- log10(peak_height)
rownames(scl_1hm) <- imputed@featureData@data$Metabolite
colnames(scl_1hm) <- pheno_data$Level
# Metabolite class color
#cols_metclass <- c("Benzenoids" = "#800000FF",
#                   "Organoheterocyclic compounds" = "#8A9045FF",
#                   "Organohalogen compounds" = "#FFA319FF",
#                   "Internal Standard" = "#767676FF")
# Add row anotation to HeatMap
#hm_row_ann <- rowAnnotation(`Class` = hm_fdata$Superclass,
#                            col = list(`Class` = cols_metclass),
#                            show_annotation_name = T,
#                            show_legend=F)
# Factor and level color
cols_factor <- c("Light conditions" = "#FFD700",
                 "Period age" = "#6E8B3D",
                 "Location" = "#00E5EE")
cols_level <- c("Alto Pano" = "#BFEFFF",
                "Alto Tena" = "#B2DFEE",
                "Talag" = "#33d2ff",
                "Light" = "#FFFF00",
                "Shade" = "#CDCD00",
                "Late" = "#9BCD9B",
                "Medium" = "#B4EEB4",
                "Early" = "#698B69")
cols_location <- c("Alto Pano" = "#BFEFFF",
                "Alto Tena" = "#B2DFEE",
                "Talag" = "#33d2ff")
cols_age <- c("Late" = "#9BCD9B",
                "Medium" = "#B4EEB4",
                "Early" = "#698B69")
cols_light <- c("Light" = "#FFFF00",
                "Shade" = "#CDCD00")

# Add top anotation to HeatMap
top_info_ann <- HeatmapAnnotation(`Factor` = pheno_data$Factor,
                                  `Level` = pheno_data$Level,
                                  col = list(`Factor` = cols_factor,
                                             `Level` = cols_level),
                                  show_annotation_name = T,
                                  show_legend=F, 
                                  border = TRUE)
# Color scale
mycol <- colorRamp2(c(3, 5, 7),
                    c("blue", "white", "red"))
#split = sample(letters[1:4], 21, replace = T)
# Heatmap matrix plotting
plot_1hm <- Heatmap(scl_1hm,
        col = mycol,
        border_gp = grid::gpar(col = "black", lty = 0.05),
        rect_gp = grid::gpar(col = "black", lwd = 0.75),
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "complete",
        top_annotation = top_info_ann,
        #column_names_gp = gpar(fontface = "italic"),
        #right_annotation = hm_row_ann,
        show_heatmap_legend = F,
        #row_split = split,
        row_km = 4,
        column_km = 2)
plot_1hm

```

Adding legends to heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Color scale legend
lgd1_1hm <- Legend(col_fun = mycol,
               title = "log10 abundance",
               at = seq(7),
               direction = "horizontal" )
# Environment factor legend
lgd2_1hm <- Legend(labels = c("Light conditions",
                          "Period age",
                          "Location"),
               legend_gp = gpar(fill = cols_factor),
               title = "Enviromental factor", ncol = 1)
lgd3_1hm <- Legend(labels = c("Alto Pano",
                          "Alto Tena",
                          "Talag"),
               legend_gp = gpar(fill = cols_location),
               title = "Location level", ncol = 1)
lgd4_1hm <- Legend(labels = c("Late",
                          "Medium",
                          "Early"),
               legend_gp = gpar(fill = cols_age),
               title = "Period age level", ncol = 1)
lgd5_1hm <- Legend(labels = c("Light",
                          "Shade"),
               legend_gp = gpar(fill = cols_light),
               title = "Light conditions level", ncol = 1)
# Metabolite class Legend
#lgd6_1hm <- Legend(labels = c(unique(hm_fdata$Superclass)) ,
#               legend_gp = gpar(fill = cols_metclass), 
#               title = "Metabolite superclass", ncol = 1)

```

ComplexHeatmap plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2500)
# Converting to ggplot
gg_1hm <- grid.grabExpr(draw(plot_1hm))
gg_1hm <- ggpubr::as_ggplot(gg_1hm)
# Legends
all_legends1 <- packLegend(lgd1_1hm, lgd2_1hm, lgd3_1hm, lgd4_1hm, lgd5_1hm, direction = "horizontal")
gg_legend1 <- grid.grabExpr(draw(all_legends1))
gg_legend1_fn <- ggpubr::as_ggplot(gg_legend1)
# Heatmap plot
lcms_1hm <- plot_grid(gg_legend1_fn,
          gg_1hm, ncol = 1,
          rel_heights = c(0.13, 0.88))
lcms_1hm
# Save heatmap plot
ggsave(filename = "Result/SubQC_HeatMap.pdf", plot = lcms_1hm,
      width = 6, height = 5, units = "in", dpi = 300, scale = 1.7)

```

## Second Heatmap plot

Here we extract and plotting heat map of samples.

Scaling, row and top heatmap anotation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2500)
# Sample data extraction
data_2hm <- no_qc[, no_qc$NewGroup != "SubGroup"]
# Extract clean data
data_2hm <- drop_flagged(data_2hm)
# Extracting feature height table
peak_2hm <- exprs(data_2hm)
# Extracting Phenotipic data
pheno_2hm <- data_2hm@phenoData@data
# Logarithmic scale
scl_2hm <- log10(peak_2hm)
rownames(scl_2hm) <- data_2hm@featureData@data$Metabolite
colnames(scl_2hm) <- pheno_2hm$Group
# Metabolite class color
#cols_metclass <- c("Benzenoids" = "#800000FF",
#                   "Organoheterocyclic compounds" = "#8A9045FF",
#                   "Organohalogen compounds" = "#FFA319FF",
#                   "Internal Standard" = "#767676FF")
# Add row anotation to HeatMap
#hm_row_ann <- rowAnnotation(`Class` = hm_fdata$Superclass,
#                            col = list(`Class` = cols_metclass),
#                            show_annotation_name = T,
#                            show_legend=F)
# Location color
cols_location <- c("Alto Pano" = "#e6550d",
                 "Alto Tena" = "#91cf60",
                 "Talag" = "#33d2ff")
# Period age color
cols_age <- c("Early" = "#FFF68F",
                 "Medium" = "#CDC673",
                 "Late" = "#8B864E")
# Light conditions color
cols_light <- c("Shade" = "#767676FF",
                 "Light" = "#FFD700")

# Add top anotation to HeatMap
top_info_ann2 <- HeatmapAnnotation(`Location` = pheno_2hm$Factor,
                                   `Light conditions` = pheno_2hm$Level,
                                   `Period age` = pheno_2hm$Age,
                                  col = list(`Location` = cols_location,
                                             `Light conditions` = cols_light,
                                             `Period age` = cols_age),
                                  show_annotation_name = T,
                                  show_legend=F, 
                                  border = TRUE)
# Color scale
mycol_hm2 <- colorRamp2(c(3, 5, 7),
                    c("blue", "white", "red"))
# Heatmap matrix plotting
plot_2hm <- Heatmap(scl_2hm,
        col = mycol_hm2,
        border_gp = grid::gpar(col = "black", lty = 0.05),
        rect_gp = grid::gpar(col = "black", lwd = 0.75),
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "complete",
        top_annotation = top_info_ann2,
        #column_names_gp = gpar(fontface = "italic"),
        #right_annotation = hm_row_ann,
        show_heatmap_legend = F,
        row_km = 3, column_km = 2)
plot_2hm

```

Adding legends to heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Color scale legend
lgd1 <- Legend(col_fun = mycol,
               title = "log10 abundance",
               at = seq(7),
               direction = "horizontal" )
# Plants species legend
lgd2 <- Legend(labels = c("Light conditions",
                          "Period age",
#                          "Sample",
                          "Location"),
               legend_gp = gpar(fill = cols_factor),
               title = "Enviromental factor", ncol = 4)
# Metabolite class Legend
#lgd3 <- Legend(labels = c(unique(hm_fdata$Superclass)) ,
#               legend_gp = gpar(fill = cols_metclass), 
#               title = "Metabolite superclass", ncol = 1)

```

ComplexHeatmap plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2500)
# Converting to ggplot
gg_heatmap <- grid.grabExpr(draw(hm_plot))
gg_heatmap <- ggpubr::as_ggplot(gg_heatmap)
# Legends
all_legends <- packLegend(lgd1, lgd2, direction = "horizontal")
gg_legend <- grid.grabExpr(draw(all_legends))
gg_legend_fn <- ggpubr::as_ggplot(gg_legend)
# Heatmap plot
gcms_hm <- plot_grid(gg_legend_fn,
          gg_heatmap, ncol = 1,
          rel_heights = c(0.195, 0.88))
gcms_hm
# Save heatmap plot
#ggsave(filename = "Result/GCMS/GCMS_Heatmap_to_DQ.pdf", plot = gcms_hm,
      #width = 5, height = 3, units = "in", dpi = 300, scale = 1.7)

```




Finish a record.

```{r}

finish_log()

```



